---
title: "Analysis of while-alive event rate for recurrent hospitalization"
author: Lu Mao
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis of while-alive event rate for recurrent hospitalization}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  \usepackage{amsmath}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## INTRODUCTION
This vignette demonstrates the use of the R-package `WA` 
in the analysis of while-alive loss (or event) rate
for recurrent event, e.g., hospitalization, in the presence of death (Mao, 2022).

### Data
Let $D$ denote the survival time and write $N_D(t)=I(D\leq t)$.
Let $N(t)$ count the number of recurrent nonfatal event 
by time $t$. Since death is a terminal event,
we must have that $N(t)=N(t\wedge D)$,
where $b\wedge c=\min(b,c)$.
Let $\mathcal H(t)=\{N_D(u), N(u):0\leq u\leq t\}$ denote
the event history up to time $t$.

### Estimands
Let $\mathcal L(\mathcal H)({\rm d}t)$ be a user-specified loss function 
that measures the instantaneous loss incurred at time $t$. 
We focus on loss functions of the form
\begin{equation}\tag{1}
\mathcal L(\mathcal H)({\rm d} t)=w_{N(t-)}{\rm d}N(t)+w^D_{N(t-)}{\rm d}N_D(t),
\end{equation}
where $w_m(t)$ and $w^D_m(t)$ are weights attached to an incident (i.e., new)
nonfatal event or death, respectively, when the patient has already experienced
$m$ nonfatal events. This scheme allows the user to assign weights 
according not only to the type and timing of the incident event but also to past history. 
It can be useful, for example, if we want to make
 the loss metric more robust to patients who experience unusually large 
numbers of event. In that case, we can specify a $w_m(t)$ that decays with larger $m$.


The cumulative loss is then defined by $\mathcal L(\mathcal H)(t)=\int_0^t L(\mathcal H)({\rm d}u)$.
 With $w_m(t)\equiv w$ and $w_m^D(t)\equiv w^D$, 
 for example, the cumulative loss becomes $\mathcal L(\mathcal H)(t)=wN(t)+w^DN_D(t)$, same as the weighted composite event process 
 considered by Mao and Lin (2016).

Let $\mathcal H^{(a)}$ denote the outcome data from group $a$
($a=1, 2,\ldots, J-1$ for different treatments and $a=0$ for the control).
To characterize the loss profile over $[0,\tau]$ in the presence of death,
define the while-alive loss rate by
\begin{equation}\tag{2}
l^{(a)}(\tau)=\frac{E\{\mathcal L(\mathcal H^{(a)})(\tau)\}}{E(D^{(a)}\wedge\tau)}.
\end{equation}
The numerator on the right hand side of (2) is 
the mean cumulative loss, akin to the mean cumulative
frequency studied by Ghosh and Lin (2000),
and the denominator
is the restricted mean survival time (RMST; Royston & Parmar, 2011),
the latter of which accounts for the length of exposure. We can interpret $l^{(a)}(\tau)$ as the 
average loss experienced per unit time alive in group $a$ over $[0, \tau]$.
Using the loss function defined in (1) with $w_m(t)\equiv 1$ and $w_m^D(t)\equiv 0$
reduces (2) to $l^{(a)}(\tau)=E\{N(\tau)\}/E(D^{(a)}\wedge t)$,
the while-alive event rate for the recurrent event considered by Schmidli et al. (2021)
and Wei et al. (2021).
The *survival-completed* cumulative loss is defined by
$L^{(a)}(\tau)=l^{(a)}(\tau)\tau$, which can be considered roughly as the total
loss over $[0,\tau]$ without being curtailed by death.


We can measure the effect size of treatment $a$ as compared to the control 
by the loss rate ratio:
\begin{equation}\tag{3}
r^{(a)}(\tau)=l^{(a)}(\tau)/l^{(0)}(\tau).
\end{equation}
That is, patients under treatment $a$ on average experience
$r^{(a)}(\tau)$ times as much loss (or as many events)
per unit time alive as compared to the control by time $\tau$.


### Tests

Testing the equality of loss rate across the $J$ groups, i.e.,
\begin{equation}\tag{4}
H_0: l^{(0)}(\tau)=l^{(1)}(\tau)=\cdots=l^{(J-1)}(\tau),
\end{equation}
is equivalent to testing $\{\log r^{(1)}(\tau),\ldots,\log r^{(J-1)}(\tau)\}^{\rm T}=\bf 0$,
which can be done via a Wald-type chi-square test with $(J-1)$ degrees of freedom (df).
Similarly, we can test 
the loss rate jointly with the RMST, i.e.,
\begin{equation}\tag{5}
H_0: l^{(0)}(\tau)=l^{(1)}(\tau)=\cdots=l^{(J-1)}(\tau)
\mbox{ and }\mu_D^{(0)}(\tau)=\mu_D^{(1)}(\tau)=\cdots=\mu_D^{(J-1)}(\tau),
\end{equation}
where $\mu_D^{(a)}(\tau)=E(D^{(a)}\wedge\tau)$,
by a Wald-type chi-square test with $2(J-1)$ df.
The $2(J-1)$ df test will be useful if $\mathcal L$ captures only
recurrent event (e.g., when $w_m^D(t)\equiv 0$ in (1)) 
so that (5) becomes a joint test of recurrent event and death. 



## BASIC SYNTAX

### Data fitting and summarization
The main function to fit the data is `LRfit()`. To use the function, the input data must be 
in the "long" format. Specifically, we need an `id` variable containing
the unique patient identifiers,  a `time` variable containing the event times,
a `status` variable labeling the event type (`status=1` for recurrent non-fatal
event, `=2` for death, and `=0` for censoring),
and, finally, a *categorical* (either binary or multiclass) `trt` variable for the treatment arm.
To analyze the (unweighted) recurrent event rate,
i.e., $w_m(t)\equiv 1$ and $w_m^D(t)\equiv 0$ in (1), simply run the function in
the default mode
```{r,eval=F}
obj<-LRfit(id,time,status,trt)
```
Alternatively, we can add the `Dweight` argument to specify a non-zero (constant) $w^D$
for death. Even more generally, we can supply self-defined $w_m(t)$ and $w_m^D(t)$
through the functional arguments `wH` and `wD`, respectively.
For example, the above code for the unweighted event rate is equivalent to
```{r,eval=F}
obj<-LRfit(id,time,status,trt,Dweight=0)
```
and to
```{r,eval=F}
# define w_m(t)
wH<-function(t,m) return (1)
# define w_m^D(t)
wD<-function(t,m) return (0)
obj<-LRfit(id,time,status,trt,wH=wH,wD=wD)
```


The returned object `obj` contains all information about the group-specific loss rate
and their standard errors. 
To extract relevant information for a particular $\tau=$`tau`, use
```{r,eval=F}
summary(obj,tau)
```
This will print out the inference results on the loss rate ratio defined in (3)
comparing each group to the reference group (which can be 
specified through the optional `ref=` argument 
if you want to override the default) 
as well as the $(J-1)$-df overall test of (4).
To request the $2(J-1)$-df loss-rate/RMST joint test of (5),
add the option `joint.test=T` to the `summary()` function.

### Plot of $L(\cdot)$
To plot the estimated survival-completed cumulative loss 
$L^{(a)}(\tau)$ as a function of $\tau$, use 
```{r,eval=F}
plot(obj,conf=TRUE)
```
This by default plots $L^{(a)}(\tau)$ for all groups.
The option `conf=T` requests the 95\% confidence limits 
for each $L^{(a)}(\tau)$ to be overlaid.
To restrict to specific groups, use
the `group=` option. 
To override the default color of the group-specific curves,
use the `group.col` argument to supply a $J$-vector of colors. Other graphical parameters 
can be added and, if so, will be passed to the underlying 
generic `plot` method.

## AN EXAMPLE WITH THE HF-ACTION TRIAL

### Data description
The Heart Failure: A Controlled Trial Investigating Outcomes of Exercise Training (HF-ACTION) study was conducted between 2003--2007 to investigate whether
 adding exercise training to the usual care of heart failure patients improves their cardiovascular outcomes (O'Conner et al., 2009). 
 We focus on a high-risk subgroup consisting of 741 nonischemic patients with 
baseline cardiopulmonary test duration less than or equal to 12 minutes
and analyze recurrent hospitalizations as well as overall survival.

Among the 741 patients, 364 were assigned to participate in exercise training sessions in addition to receiving usual care, and the remaining 377 received usual care alone as control. Over a median follow-up of 2.5 years, 49 (13.5\%) patients died in the exercise training arm, with an average of 1.8 hospitalizations per patient; 75 (19.9\%) patients died in the usual care arm, with an average of 2.0 hospitalizations per patient.
While patients undergoing exercise training experience only moderately
fewer hospitalizations in the course of follow-up, they are also much less likely to die than those in the control group. The differential survivorship needs to be taken into account when assessing the treatment effect on hospitalization. 

The associated dataset `hfaction_cpx12` is contained in the `WA` package and can be
loaded by
```{r setup}
library(WA)
head(hfaction_cpx12)
```
The dataset is already in a format suitable for `LRfit()` (`status`= 1 for hospitalization and = 2 for death).
The `time` variable is in units of years and `trt=0` for usual care (control)
and `1` for exercise training.

### Graphical analysis
First, consider the unweighted while-alive hospitalization rate.
```{r}
dat<-hfaction_cpx12
obj<-LRfit(dat$id,dat$time,dat$status,dat$trt)
## print some descriptive information 
obj
```
Let's plot the estimated group-specific survival-completed cumulative frequency.
```{r,fig.align='center',fig.width=7.2,fig.height=4.5}
plot(obj,conf=T,xlab="Time (years)",xlim=c(0, 3.5),ylim=c(0,3))
```
This pretty much reproduces the second panel in Figure 3 of Mao (2022).
We can see that the curve in the exercise training arm is substantially 
lower than that in the usual care arm, with only slightly overlapping 95\% confidence
intervals at $\tau=3.5$ years. This bodes well for the statistical significance
of treatment effect on the while-alive
hospitalization rate (recall that $l^{(a)}(\tau)=L^{(a)}(\tau)/\tau$).

### Estimation and inference
Formal analysis results at $\tau=3.5$ years can be obtained through
```{r}
## summarize the inference results at tau=3.5 years
summary(obj,tau=3.5,joint.test=T)
```
Besides the function call, the first part of the output is about
the log-loss rate: $\log l^{(0)}(\tau)$ (Ref) and
$\log r^{(1)}(\tau)=\log l^{(1)}(\tau)-\log l^{(0)}(\tau)$ (Group 1 vs 0).
We can see that the two-sided $z$-test on the log-loss rate ratio generates a 
$p$-value of 0.028, which, of course, agrees 
with the $p$-value of the 1-df chi-square test of (4)
since there are only two groups.
Further below is a table for the point and interval estimates for
the loss rate ratio: 0.830 with 95\% confidence interval
(0.703, 0.980) --- exercise training on average reduces hospitalization
by 1-0.830=17\% per year alive as compared to usual care by year 3.5.
As a result of specifying `joint.test=T`, 
we get the inference results for the RMST and the joint test of (5)
at the end. The 2-df joint test
gives a $p$-value of 0.007, suggesting that the (beneficial) treatment
effect is highly significant on survival and hospitalization jointly.

As supplemental analysis, we consider a weighted composite loss in (1)
with $w_m(t)\equiv 1$ and $w_m^D(t)\equiv 2$. This is implemented by
```{r}
## fit the data using the weighted composite loss
obj2<-LRfit(dat$id,dat$time,dat$status,dat$trt,Dweight=2)
## summarize the results at tau=3.5 years
summary(obj2,tau=3.5)
```
This shows a significant treatment effect on the
weighted composite event rate ($p$-value 0.013), consistent with the conclusion
drawn from the earlier analysis.


## References
* Ghosh, D. & Lin, D. Y. (2000). Nonparametric analysis of recurrent events and death. *Biometrics*, 56, 554--562.
* Mao, L. (2022). Nonparametric inference of general while-alive estimands for recurrent event. *Submitted*.
* Mao, L. & Lin, D. Y. (2016). Semiparametric regression for the weighted composite endpoint of recurrent and terminal events. *Biostatistics*, 172, 390--403.
* O'Connor, C. M., Whellan, D. J., Lee, K. L., Keteyian, S. J., Cooper, L. S., Ellis, S. J., ... & Rendall, D. S. (2009). Efficacy and safety of exercise training in patients with chronic heart failure: HF-ACTION randomized controlled trial. *JAMA*, 301, 1439--1450.
* Royston, P. & Parmar, M. K. (2011). The use of restricted mean survival time to estimate the treatment effect in randomized clinical trials when the proportional hazards assumption is in doubt. *Statistics in Medicine*, 30, 2409--2421.

* Schmidli, H., Roger, J. H. & Akacha, M. (2021). Estimands for recurrent event endpoints in the presence of a terminal event. 
*Statistics in Biopharmaceutical Research*, 10.1080/19466315.2021.1895883.

* Wei, J., Mutze, T., Jahn-Eimermacher, A., Roger, J. & Recurrent Event Qualification Opinion Consortium. (2021). Properties of two while-alive estimands for recurrent events and their potential estimators. *Statistics in Biopharmaceutical Research*, 10.1080/19466315.2021.1994457.

